---
title: "MATH 216 Homework 1"
author: "Alden Cowap"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    smooth_scroll: false
---

```{r, echo=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
# For read_csv() command, which is an improved version of base R's read.csv()
library(readr) 

# Load data sets. Note this assumes this .Rmd files is in the same directory as
# the .csv files.
flights <- read_csv("data/flights.csv") %>% 
  mutate(date=as.Date(date))
weather <- read_csv("data/weather.csv") %>% 
  mutate(date=as.Date(date))
planes <- read_csv("data/planes.csv")
airports <- read_csv("data/airports.csv")
states <- read_csv("data/states.csv")
```





## Admistrative:

Please indicate

* Who you collaborated with:Emily Goins
* Roughly how much time you spent on this HW so far: 5 hours
* The URL of the RPubs published URL [here](http://rpubs.com/acowap/HW1_final).
* What gave you the most trouble:Getting the unique values of flight, and uploading to Git.
* Any comments you have:





## Question 1:

Plot a "time series" of the proportion of flights that were delayed by > 30 minutes on each day.  i.e.
 
* the x-axis should be some notion of time
* the y-axis should be the proportion.

Using this plot, indicate describe the
[seasonality](https://en.wikipedia.org/wiki/Seasonality) of when delays over 30
minutes tend to occur.


```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}
library(tidyr)
flights <- flights %>% 
  mutate(delayed = ifelse(dep_delay > 30, 1, 0)) %>% 
  mutate(month_name = format(date, "%b")) %>% 
  mutate(month = as.numeric(format(date, "%m")))

flights_delay <- flights %>% 
  group_by(date) %>% 
  filter(!is.na(delayed)) %>% 
  summarise(prop = mean(delayed))

delay2 <- flights %>% select(date, month, month_name, carrier)

delay_graph <- left_join(flights_delay, delay2, by="date") %>%  
  mutate(season = ifelse(month == 12 | month == 1 | month == 2, "winter", 
                         ifelse(month == 3 | month == 4 | month == 5, "spring", 
                                ifelse(month == 6 | month == 7 | month ==8, "summer", "fall"))))

ggplot(data = delay_graph, aes(x = date, y = prop*100, color = season)) + geom_point() +
  geom_smooth(data = subset(delay_graph, month < 12), se = FALSE) + 
  labs(x = "Date", y = "Percentage of Flights Delayed", 
       title = "Seasonality of Delayed Flights")
```


## Question 2:

Some people prefer flying on older planes.  Even though they aren't as nice,
they tend to have more room.  Which airlines should these people favor?

```{r, echo=FALSE, fig.width=12, fig.height=6}

carrier = c("AA", "AS", "B6", "CO", "DL", "EV", "F9", "FL", "MQ", "OO", "UA", "US", "WN", "XE", "YV")
airline = c("American Airlines", "Alaska Airlines", "Jet Blue", "Cobaltair LTD", "Delta", "Express Jet", "Frontier Airlines", "AirTran", "Envoy Air", "SkyWest", "United Airlines", "US Airways", "SouthWest", "Deluxe Public Charter", "Mesa Airlines")

flight_code <- data.frame(carrier, airline)

planes_old <- full_join(planes, flights) %>% 
  mutate(old = ifelse(year < 2000, "before2000", "after2000")) %>% group_by(carrier) 

planes_old2 <- full_join(planes_old, flight_code) %>% 
  select(carrier, year, plane, old, airline) %>% 
  unique() %>% 
  filter(!is.na(year)) 

ggplot(data = planes_old2, aes(x=reorder(airline, year), y=2016-year)) + 
  geom_boxplot() + 
  labs(x = "Airline", y = "Average Age of Plane", title = "Age of Planes by Carrier") +
  theme(axis.text.x = element_text(angle = 35, size = 8))

```

Key for airline names:





## Question 3:

* What states did Southwest Airlines' **flight paths** tend to fly to?
* What states did Southwest Airlines' **flights** tend to fly to?

For example, Southwest Airlines Flight 60 to Dallas consists of a single flight
path, but since it flew 299 times in 2013, it would be counted as 299 flights.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}

SW <- full_join(flights, airports, by= c("dest" = "iata")) %>% filter(carrier == "WN")

#flight paths are individual observations such that flight, destination and state are all the same
SW_path <- SW %>% 
  select(flight, dest, state) %>% 
  unique() %>% 
  group_by(state) %>% 
  tally(sort = TRUE) %>% 
  filter(!is.na(state)) %>% 
  arrange(desc(n))

SW_flight <- SW %>% 
  group_by(state) %>% 
  tally(sort = TRUE) %>% 
  filter(!is.na(state)) %>% 
  arrange(desc(n))

ggplot(data = SW_path, aes(x=reorder(state, n), y = n)) + geom_bar(stat = "identity") + 
  labs(x = "Destination State", y = "Number of Flight Paths", 
       title = "SouthWest Flight Paths from Houston")

ggplot(data = SW_flight, aes(x=reorder(state, n), y = n)) + geom_bar(stat = "identity") +
  labs(x = "State", y = "Number of Flights", title = "SouthWest Flights from Houston")
```





## Question 4:

I want to know proportionately what regions (NE, south, west, midwest) each 
carrier flies to/from Houston in the month of July.  Consider the `month()`
function from the `lubridate` package.

If you're flying to the North East from Houston, you are most likely to find a flight with B6, which is Jet Blue.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}
flights_dest <- full_join(flights, airports, by = c("dest" = "iata")) 
flights_july <- full_join(flights_dest, states) %>% filter(month == 7) 

flights_july <- full_join(flights_july, flight_code) %>% 
  group_by(region, airline) %>% 
  tally(sort = TRUE) %>% 
  filter(!is.na(region))

ggplot(data = flights_july, aes(x = airline, y = n, fill = region)) + 
  geom_bar(stat = "identity", position = "fill") + 
  labs(x= "Airline", y = "Percentage of Flights in July",
         title = "Destination Regions from Houston in July") +
  theme(axis.text.x = element_text(angle = 40, size = 8))

```
